\chapter{Blocks 1--2: Unix Foundations}\label{chap:unix}

\textbf{Primary:} Command line, text processing, bash scripting\quad|\quad\textbf{Interleave:} Git, project setup

\begin{brainbox}[Why this matters]
Most bioinformatics pipelines run on Unix, and most genomic file formats are plain text. If you can wrangle text in a terminal, you can wrangle genomes.
\end{brainbox}

\begin{tipbox}[Need to install something?]
Every non-default tool named anywhere in this curriculum has install commands (macOS and Linux) and an official link in \textbf{\hyperref[app:tools]{Appendix~A}}. Flip to it the moment you hit a command that is not on your machine. Do not guess.
\end{tipbox}


\section{Block 1: The Command Line}

\sprint{1}{Navigate and Look Around}{25}
\begin{itemize}[nosep]
  \item Open your terminal. Run each command and read its output before moving on: \texttt{pwd} (print working directory), \texttt{ls} (list), \texttt{ls -la} (long format, including hidden files), \texttt{cd \textasciitilde} (go home), \texttt{cd /} (go to filesystem root), \texttt{cd -} (go to previous directory).
  \item Install \texttt{tree} if \texttt{tree --version} fails (\hyperref[app:tools]{Appendix~A}), then run \texttt{tree \textasciitilde} with no arguments.
  \item Make a project directory: \texttt{mkdir -p \textasciitilde/bioinfo/block1/\{data,scripts,results\}}. The curly braces are a \textit{shell brace expansion} --- the shell rewrites this into three \texttt{mkdir -p} arguments before running the command.
  \item Explore: what is in \texttt{/usr/bin}? How many entries? (\texttt{ls /usr/bin | wc -l})
\end{itemize}
\begin{tipbox}[Getting unstuck at the shell]
\texttt{Tab} completes filenames and commands. \texttt{Up-arrow} brings back the previous command. \texttt{Ctrl-C} cancels whatever is running. \texttt{Ctrl-L} (or \texttt{clear}) clears the screen. \texttt{man ls} opens the manual for a command; press \texttt{q} to quit. If you ever see a stray \texttt{>} prompt you did not expect, you left an open quote; press \texttt{Ctrl-C} and start over.
\end{tipbox}
\begin{winbox}[Checkpoint]
Run \texttt{tree \textasciitilde/bioinfo} and paste its output into \texttt{journal.md} --- you should see the three empty subdirectories you just made.
\end{winbox}

\sprint{2}{Pipes: Chaining Small Tools}{25}

\begin{warnbox}[Read this before you paste any long command from this curriculum]
Several commands below are too long to fit on one printed line, so either the source splits them across lines (with backslashes or pipes) or the PDF wraps them visually. Two rules that will save you hours:
\begin{itemize}[nosep]
  \item \textbf{A backslash (\textbackslash) at the end of a line means ``continue this command on the next line.''} The backslash must be the \textit{very last} character; a single trailing space turns \texttt{\textbackslash~} into ``escape a space'' and the line continuation is lost. Do not put anything (not even a \texttt{\#} comment) after the backslash.
  \item \textbf{Inside a URL or any other token, \texttt{\textbackslash}-then-newline is consumed with no replacement character.} So \texttt{https://host/\textbackslash}$\hookleftarrow$\texttt{path} reassembles to \texttt{https://host/path} with no space inserted. This is why you will see URLs broken across lines below with no leading whitespace on the continuation.
\end{itemize}
A safer alternative when you have a choice: end the line on a pipe (\texttt{|}) or a logical operator (\texttt{\&\&}, \texttt{||}) instead. The shell knows a pipeline is incomplete, so it reads the next line automatically; no backslash is needed. If a copy-paste from the PDF misbehaves, suspect a smart-quote, a stripped backslash, or a soft line break --- retype the break by hand.
\end{warnbox}

\begin{itemize}[nosep]
  \item The pipe character \texttt{|} sends the standard output of one command into the standard input of the next. This is the core of the ``small tools, composed'' Unix design.
  \item Try: \texttt{echo 'ATGCGATCGATCG' | grep -o . | sort | uniq -c | sort -rn}. Read it left to right: print the string, split into one character per line, sort the lines, count adjacent duplicates, sort by count descending. You should see A, C, G, T each with a count.
  \item Download the S288C yeast reference genome (tarball is $\sim$4~MB). The URL is long, so the source below splits it with backslash continuation --- the continuation lines start at column 1 with no leading spaces so the URL reassembles cleanly:
    \begin{lstlisting}
curl -O \
https://downloads.yeastgenome.org/sequence/\
S288C_reference/genome_releases/\
S288C_reference_genome_Current_Release.tgz
tar xzf S288C_reference_genome_Current_Release.tgz
cd S288C_reference_genome_*
    \end{lstlisting}
  \item Find the genome FASTA: \texttt{ls *.fsa}. Look at the header lines with \texttt{grep '>' *.fsa | head}. Each header line starts with \texttt{>} and names one chromosome (I through XVI) plus the mitochondrial genome --- that is what \texttt{grep '>'} pulls out.
\end{itemize}
\begin{tipbox}[If \texttt{curl} fails or the URL has changed]
Try \texttt{wget} instead of \texttt{curl -O}, or open \url{https://www.yeastgenome.org/} and navigate to Sequence $\to$ Download. A long error with ``Could not resolve host'' usually means no network; ``404 Not Found'' means the file moved.
\end{tipbox}
\begin{winbox}[Checkpoint]
You chained four commands with pipes, counted nucleotide frequencies, and downloaded and un-tarred the yeast reference. You also inspected the FASTA and identified what each header line names.
\end{winbox}

\sprint{3}{Grep and Awk: Extracting Lines and Columns}{25}
\begin{itemize}[nosep]
  \item \texttt{grep} searches for patterns. \texttt{grep -c '>' file.fasta} counts sequences; \texttt{grep -v} inverts the match (keeps lines that do \textit{not} match).
  \item \texttt{awk} extracts and transforms whitespace- or tab-separated columns. \texttt{awk -F'\textbackslash t' '\{print \$1, \$2, \$5\}' file.vcf} prints columns 1, 2, and 5 of a tab-delimited file --- for a VCF those are chromosome, position, and quality.
  \item Download a small yeast VCF to practice on. Same backslash-continuation pattern as the S288C download in Sprint 2:
    \begin{lstlisting}
curl -O \
https://ftp.ensembl.org/pub/current_variation/\
vcf/saccharomyces_cerevisiae/\
saccharomyces_cerevisiae.vcf.gz
gunzip saccharomyces_cerevisiae.vcf.gz
    \end{lstlisting}
    If the link has moved, browse \url{https://ftp.ensembl.org/pub/current_variation/vcf/} and pick any small organism's VCF.
  \item Look at the file structure: \texttt{head -n 30 saccharomyces\_cerevisiae.vcf}. Lines starting with \texttt{\#\#} are metadata; the single line starting with \texttt{\#CHROM} is the column header; all other lines are data records.
  \item Count variants per chromosome. Here the line endings are pipes, so the shell knows the pipeline is incomplete and no backslashes are needed:
    \begin{lstlisting}
grep -v '^#' saccharomyces_cerevisiae.vcf |
  awk '{print $1}' |
  sort | uniq -c | sort -rn
    \end{lstlisting}
    \texttt{grep -v '\textasciicircum\#'} drops every header line; \texttt{awk} pulls out the chromosome column; \texttt{sort | uniq -c} tallies each chromosome; the final \texttt{sort -rn} orders by count.
\end{itemize}
\begin{winbox}[Checkpoint]
Running the pipeline above produces a table with roughly 16 rows (one per yeast chromosome) with a count in front of each. Paste this table into \texttt{journal.md}.
\end{winbox}

\sprint{4}{Put It All Together: Pick a Restriction Enzyme for Cloning}{25}
A real scenario: you want to shotgun-clone roughly 3~kb fragments of the yeast genome into a plasmid. Three cheap 6-cutters are on the shelf: EcoRI (\texttt{GAATTC}), BamHI (\texttt{GGATCC}), and HindIII (\texttt{AAGCTT}). Before ordering any of them, predict from the genome sequence how often each enzyme cuts, and which one produces fragments closest to your 3~kb target. Do it all with pipes; no scripts yet. Paste every command and its output into \texttt{journal.md} as you go.
\begin{enumerate}[nosep]
  \item \textbf{Confirm and organize.} \texttt{head} the yeast FASTA --- the first line should start with \texttt{>}. Then move the FASTA (and the VCF from Sprint 3) into \texttt{data/} in one pipeline: \texttt{ls *.fsa *.vcf 2>/dev/null | xargs -I\{\} mv \{\} \textasciitilde/bioinfo/block1/data/}. Verify with \texttt{ls \textasciitilde/bioinfo/block1/data/}.
  \item \textbf{Measure the genome.} Strip headers and newlines, count characters:
    \begin{lstlisting}
grep -v '>' ~/bioinfo/block1/data/*.fsa | \
  tr -d '\n' | wc -c
    \end{lstlisting}
    For S288C, expect roughly 12\,Mb. Save this number --- you'll divide by it next.
  \item \textbf{Count cut sites per enzyme.} Template for EcoRI:
    \begin{lstlisting}
grep -v '>' ~/bioinfo/block1/data/*.fsa | \
  tr -d '\n' | grep -o GAATTC | wc -l
    \end{lstlisting}
    Run it three times --- once each for \texttt{GAATTC}, \texttt{GGATCC}, \texttt{AAGCTT}. Which cuts most? Fewest?
  \item \textbf{Why one strand is enough.} All three recognition sites are palindromes (they read the same on both strands), so every real cut site is already captured by a forward-strand match. For non-palindromic motifs you would also have to scan the reverse complement; good to know before you generalize this approach.
  \item \textbf{Compute average fragment size.} Average fragment size $\approx$ genome length $\div$ number of cut sites. Use \texttt{bc}: \texttt{echo "12157105 / NSITES" | bc} (swap in your actual genome length and cut count). Expect numbers in the range of hundreds of bp to tens of kb.
  \item \textbf{Pick your enzyme.} Which enzyme's average fragment size is closest to your 3~kb target? Write one sentence in \texttt{journal.md} explaining your choice, and flag which enzyme would be a bad idea (cuts so often you would get tiny fragments, or so rarely the fragments are too big to clone).
\end{enumerate}
\begin{winbox}[Checkpoint]
Your journal shows the \texttt{head} check, the move-into-\texttt{data/} pipeline, the genome length, cut-site counts for three enzymes, their average fragment sizes, and your enzyme pick with a one-sentence justification, all produced with pipes.
\end{winbox}

\begin{tipbox}[Pattern to remember]
In Unix bioinformatics, the workflow is almost always: extract columns (\texttt{cut} / \texttt{awk}) $\to$ filter rows (\texttt{grep}) $\to$ sort $\to$ count (\texttt{uniq -c}). This pattern covers a large fraction of quick data exploration.
\end{tipbox}


\section{Block 2: Scripts, Environments \& Git}

\sprint{5}{Your First Bash Script}{25}
\begin{itemize}[nosep]
  \item Create the file: \texttt{nano \textasciitilde/bioinfo/block1/scripts/count\_variants.sh} (or use VS Code --- any plain-text editor is fine).
  \item The first line must be the ``shebang'' \texttt{\#!/usr/bin/env bash}, which tells the OS which interpreter to run. Follow it with \texttt{set -euo pipefail}: \texttt{-e} exits on the first error, \texttt{-u} errors on undefined variables, \texttt{-o pipefail} fails if any command in a pipeline fails.
  \item Write a script that reads a VCF filename from its first argument (\texttt{\$1}) and prints a per-chromosome variant count. Template to fill in:
    \begin{lstlisting}
#!/usr/bin/env bash
set -euo pipefail

vcf="${1:?usage: count_variants.sh FILE.vcf}"
grep -v '^#' "$vcf" | awk '{print $1}' | \
  sort | uniq -c | sort -rn
    \end{lstlisting}
    The \texttt{\$\{1:?\ldots\}} syntax prints a usage message and exits if you forget to pass an argument.
  \item Make it executable and run it: \texttt{chmod +x count\_variants.sh} then \texttt{./count\_variants.sh \textasciitilde/bioinfo/block1/data/saccharomyces\_cerevisiae.vcf}.
\end{itemize}
\begin{winbox}[Checkpoint]
The script prints the same per-chromosome counts you computed by hand in Sprint 3. Run it on a different VCF (same script, different data) and confirm it still works.
\end{winbox}

\sprint{6}{Conda and Mamba: Isolated, Reproducible Environments}{25}
\begin{itemize}[nosep]
  \item Create an environment: \texttt{mamba create -n biotools -c conda-forge -c bioconda samtools bcftools bedtools fastp}. The \texttt{-c} flags set channels; put \texttt{conda-forge} before \texttt{bioconda}.
  \item Activate it: \texttt{mamba activate biotools}. Verify: \texttt{samtools --version}.
  \item Export it: \texttt{mamba env export > environment.yml}. Anyone can then recreate your setup with \texttt{mamba env create -f environment.yml}.
  \item Deactivate: \texttt{mamba deactivate}. Run \texttt{samtools} again and confirm it is no longer on \texttt{PATH}. Environments isolate dependencies from each other and from the system.
  \item \textbf{Why mamba?} Conda's own dependency solver is slow on large environments. Mamba is a drop-in replacement (written in C++) that resolves the same specifications in seconds. Miniforge (\url{https://github.com/conda-forge/miniforge}) ships with it. You can still type \texttt{conda} everywhere; \texttt{mamba} is just faster.
\end{itemize}
\begin{tipbox}[Common first-time conda mistakes]
\begin{itemize}[nosep]
  \item Forgetting \texttt{-c bioconda} when installing bioinformatics packages. You will get a \texttt{PackagesNotFoundError}.
  \item Installing into the \texttt{base} environment. Do not: \texttt{base} is for conda itself. Always make a named environment.
  \item Mixing \texttt{pip install} with \texttt{conda install} in the same environment without care. If you must, run \texttt{pip} last.
  \item Forgetting to \texttt{mamba activate}. A freshly opened terminal has no environment active; your \texttt{biotools} tools will look missing.
\end{itemize}
\end{tipbox}
\begin{warnbox}[Cross-architecture caveats]
Conda packages are built per (OS, CPU) pair --- \texttt{linux-64}, \texttt{osx-64}, \texttt{osx-arm64} (Apple Silicon), \texttt{linux-aarch64} (ARM Linux / some HPCs). Not every bioinformatics package has a build for every platform. On an M1/M2/M3 Mac, a missing \texttt{osx-arm64} build usually surfaces as \texttt{PackagesNotFoundError}; common fixes are to (1) create the env under Rosetta with \texttt{CONDA\_SUBDIR=osx-64 conda create \ldots}, or (2) find an alternative package with native ARM support. An \texttt{environment.yml} built on one platform is not guaranteed to solve on another --- expect small fix-ups when moving between laptop and cluster, and note the architecture in a comment at the top of the file.
\end{warnbox}
\begin{winbox}[Checkpoint]
Your \texttt{environment.yml} is a reproducible recipe. Commit it to Git, with a header comment noting which architecture built it.
\end{winbox}

\sprint{7}{Containers: Reproducibility Beyond Conda}{25}
\begin{brainbox}[What is a container?]
A container is a single image file that bundles an entire Linux userland (OS libraries, the tool you care about, and everything that tool depends on) so that it runs identically on any compatible host. Think of the software stack as a stack of layers: the kernel at the bottom, system libraries on top of the kernel, language runtimes (Python, R) on top of those, and your scripts at the very top. Conda manages the \textit{upper} layers: Python and R packages, plus a handful of binaries that link against them. Containers freeze \textit{lower} layers too: the system libraries, \texttt{glibc} version, and compiled binaries that conda normally inherits from whatever host it runs on. A container is not a virtual machine (there is no guest kernel), so startup is near-instant and overhead is minimal.

Rule of thumb: use conda for interactive development (fast to edit, easy to iterate). Use a container when you ship a pipeline to a cluster, share an analysis with a collaborator on a different OS, or archive a workflow alongside a manuscript --- situations where ``works on my machine'' has to become ``works on yours.''
\end{brainbox}

\begin{itemize}[nosep]
  \item \textbf{Install Docker.} Docker is the developer-facing runtime. Install Docker Desktop on macOS (\url{https://docs.docker.com/desktop/install/mac-install/}) or Docker Engine on Linux (\url{https://docs.docker.com/engine/install/}). Verify with \texttt{docker --version}.
  \item \textbf{Pull a biocontainer.} The Biocontainers project auto-builds one container per bioconda package. Pull bcftools, which we'll use on the VCF you already have from Sprint 3:
    \begin{lstlisting}
docker pull \
  quay.io/biocontainers/bcftools:1.19--h8b25389_0
    \end{lstlisting}
  \item \textbf{Run without installing.} Execute the container's \texttt{bcftools} with no other setup:
    \begin{lstlisting}
docker run --rm \
  quay.io/biocontainers/bcftools:1.19--h8b25389_0 \
  bcftools --version
    \end{lstlisting}
    \texttt{--rm} discards the stopped container after it exits; the image stays cached for next time.
  \item \textbf{Mount your data.} Containers see an isolated filesystem by default. Map a host directory in so the tool can read local files. \texttt{cd} into the directory holding your VCF first, then:
    \begin{lstlisting}
cd ~/bioinfo/block1/data
docker run --rm -v "$PWD":/work -w /work \
  quay.io/biocontainers/bcftools:1.19--h8b25389_0 \
  bcftools view -h saccharomyces_cerevisiae.vcf | head -20
    \end{lstlisting}
    \texttt{-v HOST:CONTAINER} mounts a host directory; \texttt{-w} sets the working directory inside. The container has no \texttt{bcftools} on the host, but it ran one anyway --- and produced output from your real VCF.
\end{itemize}
\begin{tipbox}[Conda vs.\ container --- when to reach for which]
Conda gives you speed and easy iteration at the cost of strict reproducibility: two users with the same \texttt{environment.yml} may still resolve to different package builds on different platforms. A container gives you a fixed binary that will run identically for years, at the cost of being heavier to build and harder to edit. Many published pipelines use both: conda during development, a container (often built from the conda env) for the frozen release.
\end{tipbox}
\begin{winbox}[Checkpoint]
You pulled a container, ran a bioinformatics tool from it without installing anything locally, and mounted a host directory into it so it could read your real VCF.
\end{winbox}

\sprint{8}{Git and GitHub: Version Control and Your Portfolio}{25}
\begin{itemize}[nosep]
  \item \textbf{Initialize and commit.}
    \begin{lstlisting}
cd ~/bioinfo && git init -b main && \
  git add -A && \
  git commit -m 'Blocks 1-2: Unix foundations'
    \end{lstlisting}
  \item Edit something. \texttt{git diff} to inspect, \texttt{git add}/\texttt{git commit} to save, \texttt{git log --oneline} to review.
  \item \textbf{.gitignore.} Exclude anything big or generated: \texttt{*.bam}, \texttt{*.fastq.gz}, \texttt{*.vcf.gz}, \texttt{*.sam}. Never commit raw data. Commit the code that produces it and a pointer to where it lives.
  \item \textbf{Set up an SSH key for GitHub} (one-time; after this, no password prompts):
    \begin{lstlisting}
ssh-keygen -t ed25519 -C "you@example.com"
# press Enter to accept the default path;
# optionally set a passphrase
cat ~/.ssh/id_ed25519.pub
    \end{lstlisting}
    Copy the printed public key, then go to github.com $\to$ Settings $\to$ SSH and GPG keys $\to$ New SSH key and paste it. Test: \texttt{ssh -T git@github.com} should greet you by username.
  \item \textbf{Publish your repo.} Create an empty repo on github.com (don't add a README --- your local copy already has files). Then from your local repo:
    \begin{lstlisting}
git remote add origin \
  git@github.com:<you>/bioinfo-journey.git
git push -u origin main
    \end{lstlisting}
    The \texttt{-u} flag wires \texttt{main} to the remote so future \texttt{git push}es need no arguments.
  \item \textbf{Clone and pull.} To download a repo onto a new machine, or to grab someone else's work: \texttt{git clone git@github.com:user/repo.git}. Inside an existing clone, \texttt{git pull} fetches new commits from the remote and merges them into your branch; that is how you keep a checkout in sync. Practice: clone your own repo into \texttt{/tmp/}, push a change from your main checkout, then run \texttt{git pull} in \texttt{/tmp/} and confirm the change arrived.
\end{itemize}
\begin{tipbox}[Common Git traps]
\begin{itemize}[nosep]
  \item \texttt{git push} refused with ``non-fast-forward'': someone (or past-you on a different machine) pushed commits you do not have. Run \texttt{git pull --rebase} first, then push.
  \item You committed a large data file by accident. Removing it with \texttt{git rm} only hides it going forward; the old commits still contain it. For a solo repo it is usually simplest to add the pattern to \texttt{.gitignore} and keep moving; for bigger accidents use \texttt{git filter-repo} or the BFG tool.
  \item \texttt{git status} shows files you did not edit. Line-ending conversion on Windows/WSL is a common culprit: set \texttt{git config --global core.autocrlf input}.
\end{itemize}
\end{tipbox}
\begin{winbox}[Checkpoint]
Your bioinfo-journey repo has real commits, ignores big files, pushes over SSH without password prompts, and you have proved you can clone and pull from it. This is the start of your portfolio.
\end{winbox}

\sprint{9}{Integration Sprint: Reproducible Variant Summary}{50}
The goal: take the work from Sprints 3--5 and turn it into a single reproducible script that runs inside the conda environment from Sprint 6, writes a timestamped report, and is tracked in the Git repo from Sprint 8. No new tools --- just the ones you've already installed.
\begin{enumerate}[nosep]
  \item \textbf{Write \texttt{scripts/variant\_report.sh}} that takes a VCF path as \texttt{\$1} and:
    \begin{itemize}[nosep]
      \item starts with \texttt{\#!/usr/bin/env bash} and \texttt{set -euo pipefail} (Sprint 5);
      \item runs the per-chromosome variant tally from Sprint 3 (\texttt{grep -v '\textasciicircum\#' | awk \ldots | sort | uniq -c | sort -rn});
      \item also runs \texttt{bcftools stats "\$1" | head -30} from the \texttt{biotools} env (Sprint 6);
      \item writes both outputs, prefixed by \texttt{date} and the input filename, into \texttt{results/report\_\$(date +\%Y\%m\%d).txt}.
    \end{itemize}
  \item \textbf{Use the conda env.} Either \texttt{mamba activate biotools} before running the script, or have the script source conda and activate the env itself (\texttt{source "\$(conda info --base)/etc/profile.d/conda.sh"; conda activate biotools}).
  \item \textbf{Run it} on the yeast VCF from Sprint 3 and confirm the report file appears under \texttt{results/}.
  \item \textbf{Commit the script and the report.} Per the \texttt{.gitignore} rules from Sprint 8, do not commit the raw VCF --- but the script and the small text report are fair game.
\end{enumerate}
\begin{winbox}[Checkpoint]
A single command,
\begin{lstlisting}
./scripts/variant_report.sh \
  ~/bioinfo/block1/data/saccharomyces_cerevisiae.vcf
\end{lstlisting}
produces a dated report under \texttt{results/} that you can hand to someone else along with your repo and your \texttt{environment.yml}, and they can reproduce it.
\end{winbox}

\begin{restbox}[End of Blocks 1--2]
Take a day to step away before starting Python. The command-line fundamentals are now in place.
\end{restbox}

\subsection{Can I move on? Checkpoint:}
\begin{itemize}[nosep]
  \item You can navigate any directory, find files, and chain 3+ commands with pipes
  \item You can write a bash script that processes a genomic file and logs output
  \item You have a conda environment with bioinformatics tools installed
  \item Your work is in a Git repo on GitHub
\end{itemize}
\textit{If not, spend 2--3 more days on the gaps. No rush.}

\subsection{Resources}
\begin{itemize}[nosep]
  \item Software Carpentry: The Unix Shell (free, interactive)
  \item Buffalo -- \textit{Bioinformatics Data Skills}, Ch.\ 3, 7, 12
  \item explainshell.com --- paste any command and see what each part does
\end{itemize}
