\chapter{Blocks 3--4: Python for Genomic Data}\label{chap:python}

\textbf{Primary:} Data structures, Pandas, visualization\quad|\quad\textbf{Interleave:} Unix pipes, real genomic files

\begin{brainbox}[Connection to what you already know]
You just learned to process text files with grep and awk. Python does the same things but with reusable code, richer data structures, and the ability to make plots and run statistics in the same script.
\end{brainbox}

\begin{tipbox}[Running Python: a newbie orientation]
You need Python 3 (not 2). Check with \texttt{python3 --version}; Miniforge's \texttt{base} environment already has a recent Python 3. Two ways to run code:
\begin{itemize}[nosep]
  \item \textbf{REPL / notebook.} Type \texttt{python3} at the shell for a live prompt, or \texttt{jupyter lab} after \texttt{mamba install -c conda-forge jupyterlab}. Good for exploration.
  \item \textbf{Script.} Put code in \texttt{file.py} and run \texttt{python3 file.py}. This is what the exercises below expect.
\end{itemize}
Indentation is syntax in Python: four spaces per level, consistent throughout a file. Mixing tabs and spaces produces \texttt{IndentationError}. Configure your editor to insert spaces when you press \texttt{Tab}.
\end{tipbox}


\section{Block 3: Core Python via Biology}

\sprint{1}{Data Structures for Biology}{25}
Pythonic code is less about clever tricks and more about reaching for the right built-in collection. Each has a natural biological use case. \textbf{Open \texttt{python3} (or a Jupyter cell) and run each block before moving on} --- read-only is not enough here.
\begin{itemize}[nosep]
  \item \textbf{Strings = DNA sequences.} Reverse-complement \texttt{ATGCGATCG} and verify by hand:
    \begin{lstlisting}
seq = 'ATGCGATCG'
comp = seq.translate(str.maketrans('ATCG','TAGC'))
print(comp[::-1])  # reverse complement
    \end{lstlisting}
    \textbf{Do:} now reverse-complement \texttt{AAACGTTTGCAGGGAATTC} and find the position of the EcoRI site (\texttt{GAATTC}) in both the original and the reverse complement using \texttt{str.find()}.
  \item \textbf{Lists = ordered, mutable collections.} Build a list of gene names and filter it:
    \begin{lstlisting}
genes = ['BRCA1', 'BRCA2', 'TP53',
         'BRAF', 'MYC', 'BRD4']
brc = [g for g in genes if g.startswith('BR')]
print(brc)  # 4 hits
    \end{lstlisting}
    \textbf{Do:} change the comprehension to keep only genes whose name is exactly four characters long. Swap the outer brackets for parentheses --- you now have a \textit{generator expression}; pass it to \texttt{list()} to get the same result.
  \item \textbf{Tuples = immutable records.} Genomic coordinates are a natural fit:
    \begin{lstlisting}
peak = ('chr1', 1000, 2000)
chrom, start, end = peak  # unpack
print(end - start)        # 1000
    \end{lstlisting}
    \textbf{Do:} put three peak tuples into a \texttt{set()}; add a duplicate of the first; print \texttt{len()}. Because tuples are hashable, the duplicate is dropped --- this is how you build ``regions I've already processed.''
  \item \textbf{Dictionaries = lookup tables.} Build a tiny codon table and translate three bases at a time:
    \begin{lstlisting}
codons = {'ATG': 'M', 'GCT': 'A',
          'AAA': 'K', 'TAA': '*'}
mrna = 'ATGGCTAAATAA'
prot = ''.join(codons[mrna[i:i+3]]
               for i in range(0, len(mrna), 3))
print(prot)  # MAK*
    \end{lstlisting}
    \textbf{Do:} change one codon in \texttt{mrna} so the protein reads \texttt{MA*}.
  \item \textbf{Sets = membership and overlap.} Compare two ``hit lists'':
    \begin{lstlisting}
sampleA = {'BRCA1', 'TP53', 'MYC'}
sampleB = {'TP53', 'MYC', 'EGFR'}
print(sampleA & sampleB)  # in both
print(sampleA - sampleB)  # only A
    \end{lstlisting}
    \textbf{Do:} compute the union with \texttt{|} and confirm its size equals \texttt{len(A) + len(B) - len(A \& B)}.
  \item \textbf{\texttt{collections.Counter} and \texttt{defaultdict}.} The right tool for ``count things'' and ``group things'':
    \begin{lstlisting}
from collections import Counter, defaultdict
seq = 'ATGCATGCATGCAA'
print(Counter(seq).most_common())

variants = [('chr1', 100), ('chr2', 50),
            ('chr1', 250), ('chr2', 80)]
by_chrom = defaultdict(list)
for chrom, pos in variants:
    by_chrom[chrom].append(pos)
print(dict(by_chrom))
    \end{lstlisting}
    \textbf{Do:} use \texttt{Counter} on a longer DNA string and call \texttt{.most\_common(2)} to get the top two bases.
\end{itemize}
\begin{winbox}[Checkpoint]
You reverse-complemented DNA, filtered a list, deduplicated peaks via a set, translated mRNA with a dict, intersected two sample hit lists, and grouped variants by chromosome with \texttt{defaultdict}. Idiomatic Python is largely about picking the right container rather than writing loops by hand.
\end{winbox}

\sprint{2}{Parsing Genomic Files}{25}
\begin{brainbox}[What is a parser?]
A \textbf{parser} is just a function that turns a text file into structured data your program can use --- a list of records, a dictionary, a dataclass. Every bioinformatics format (FASTA, FASTQ, VCF, BED, GFF) is plain text, but each has its own line-level rules: which lines are headers, which are data, what separates fields. Writing a parser is the act of teaching Python those rules. The recipe is always the same: open the file, walk through lines one at a time, recognize what each line means, and accumulate records. Today we'll build that recipe up step by step.
\end{brainbox}

\begin{itemize}[nosep]
  \item \textbf{Step 1: read a file, line by line.} Make a tiny test file in your current directory called \texttt{toy.tsv}:
    \begin{lstlisting}
chr1    100    200
chr1    500    600
chr2    50     75
    \end{lstlisting}
    Then in Python:
    \begin{lstlisting}
with open('toy.tsv') as f:
    for line in f:
        print(repr(line))
    \end{lstlisting}
    The \texttt{with} block is a \textit{context manager} --- it closes the file automatically even if an exception is raised. \texttt{repr()} reveals the trailing \texttt{\textbackslash n} on each line.
  \item \textbf{Step 2: split each line into fields.} Replace the \texttt{print} with:
    \begin{lstlisting}
chrom, start, end = line.strip().split('\t')
print(chrom, int(start), int(end))
    \end{lstlisting}
    \texttt{strip()} drops the trailing newline; \texttt{split('\textbackslash t')} gives a list of three strings; \texttt{int()} converts the numeric fields. You just wrote a one-line BED parser.
  \item \textbf{Step 3: return structured records.} Wrap it in a function that returns a list of dictionaries:
    \begin{lstlisting}
def parse_bed(path):
    records = []
    with open(path) as f:
        for line in f:
            chrom, start, end = (
                line.strip().split('\t'))
            records.append({
                'chrom': chrom,
                'start': int(start),
                'end':   int(end),
            })
    return records

print(parse_bed('toy.tsv'))
    \end{lstlisting}
    BED is tab-separated, 0-based, half-open --- those are format conventions, not Python rules.
  \item \textbf{Step 4: now FASTA --- one wrinkle.} A FASTA \textit{record} is a header line starting with \texttt{>} followed by one or more sequence lines that you have to glue together. So unlike BED, you can't return a record from each line: you accumulate sequence lines until you see the next \texttt{>}. Write \texttt{toy.fa}:
    \begin{lstlisting}
>seq1 short
ATGCATGC
ATGC
>seq2 even shorter
GGGTTT
    \end{lstlisting}
    Then write \texttt{parse\_fasta(path)} that returns a list of \texttt{(header, sequence)} tuples. Sketch:
    \begin{lstlisting}
def parse_fasta(path):
    records = []
    header, chunks = None, []
    with open(path) as f:
        for line in f:
            line = line.rstrip()
            if line.startswith('>'):
                if header is not None:
                    records.append(
                        (header, ''.join(chunks)))
                header = line[1:]   # drop '>'
                chunks = []
            else:
                chunks.append(line)
        if header is not None:       # last record
            records.append(
                (header, ''.join(chunks)))
    return records
    \end{lstlisting}
    Run \texttt{parse\_fasta('toy.fa')} and confirm \texttt{seq1}'s sequence is \texttt{ATGCATGCATGC} (the two lines joined).
  \item \textbf{Step 5: handle malformed input.} Add a deliberately broken line to \texttt{toy.tsv} (e.g.\ remove a tab) and rerun \texttt{parse\_bed()}. You'll get a \texttt{ValueError}. Wrap the inner work in \texttt{try/except ValueError}: print a warning naming the line, then \texttt{continue}. A robust parser tolerates one bad row without crashing the whole run.
\end{itemize}
\begin{winbox}[Checkpoint]
You wrote two parsers from scratch: a single-line BED parser and a multi-line FASTA parser. Both run on tiny test files you wrote by hand. Your BED parser also prints a warning and keeps going when you feed it a malformed line.
\end{winbox}

\sprint{3}{Generators: Streaming Files Larger Than RAM}{25}
\begin{itemize}[nosep]
  \item A generator function uses \texttt{yield} instead of \texttt{return}; it produces items one at a time without building the full list in memory.
  \item Convert your FASTA parser to a generator. Usage changes from \texttt{records = parse(path)} followed by \texttt{for r in records} to a single \texttt{for r in parse(path)}. The file stays open only as long as the loop runs.
  \item Why this matters: a whole-genome VCF or a raw FASTQ can be tens of GB. Loading all of it into memory will swap your machine or crash the job.
\end{itemize}
\begin{tipbox}[One detail newbies miss]
A generator is consumed once. After \texttt{for r in parse(path): \ldots}, the generator is exhausted; running the loop again yields nothing. If you need the data twice, either call \texttt{parse(path)} again (re-opens the file) or materialize with \texttt{list(parse(path))}.
\end{tipbox}
\begin{winbox}[Checkpoint]
Your FASTA parser yields one record at a time and runs on a 1~MB test file without loading all of it into memory. Verify with \texttt{import sys; sys.getsizeof(parse(path))} before and after converting.
\end{winbox}

\sprint{4}{Pythonic Idioms and Type Hints}{25}
Writing Python that other people (and future you) can read quickly is mostly a matter of using the idioms the language was built around. Each of these takes a few minutes to learn and saves hours of debugging.
\begin{itemize}[nosep]
  \item \textbf{f-strings.} \texttt{f"\{gene\} on chr\{chrom\} (padj=\{pval:.2e\})"} --- variables, formatting, and expressions in one readable line. Use these exclusively for user-facing strings; they're faster than \texttt{\%} or \texttt{.format()} and much harder to misread.
  \item \textbf{\texttt{pathlib} for filesystem paths.}
    \begin{lstlisting}
from pathlib import Path
data = Path.home() / "bioinfo" / "block1" / "data"
for vcf in data.glob("*.vcf"):
    print(vcf.stem, vcf.stat().st_size)
    \end{lstlisting}
    Path objects beat string concatenation: they handle OS differences, support \texttt{/} for joining, and expose \texttt{.stem}, \texttt{.suffix}, \texttt{.parent}, \texttt{.exists()}.
  \item \textbf{\texttt{enumerate} and \texttt{zip}.} Almost never write \texttt{for i in range(len(xs))} --- use \texttt{for i, x in enumerate(xs)} or \texttt{for a, b in zip(xs, ys)}. These are the canonical ways to walk sequences.
  \item \textbf{Dataclasses for records.}
    \begin{lstlisting}
from dataclasses import dataclass

@dataclass
class Variant:
    chrom: str
    pos: int
    ref: str
    alt: str
    qual: float = 0.0
    \end{lstlisting}
    You get \texttt{\_\_init\_\_}, \texttt{\_\_repr\_\_}, and equality for free, with attribute access (\texttt{v.pos}) instead of \texttt{v["pos"]}. Prefer these over parallel lists or dicts-of-fields for anything with a fixed shape.
  \item \textbf{Type hints.} Annotations like \texttt{def parse\_vcf(path: Path) -> list[Variant]:} are not enforced at runtime, but tools (\texttt{mypy}, Pyright, your editor) use them to catch bugs before you run the code, and they're the best function documentation you can write. Start with hints on function signatures and dataclass fields --- don't try to annotate everything at once.
  \item \textbf{Context managers beyond \texttt{open()}.} \texttt{with pysam.AlignmentFile(\ldots) as bam:}, \texttt{with tempfile.TemporaryDirectory() as d:}, \texttt{with contextlib.suppress(FileNotFoundError):}. Any resource that needs deterministic cleanup --- file handles, temp dirs, database cursors, locks --- is a candidate.
\end{itemize}
\begin{tipbox}[Reach for the stdlib first]
Before adding a dependency, check \texttt{collections}, \texttt{itertools}, \texttt{functools}, \texttt{pathlib}, \texttt{dataclasses}, \texttt{contextlib}, \texttt{statistics}. The Python standard library solves most problems you'll hit in genomics scripting, and stdlib code ages well --- no surprise breaking changes when your analysis reviewer tries to rerun it two years later.
\end{tipbox}
\begin{winbox}[Checkpoint]
You wrote a function with a type-hinted signature, a dataclass for one biological record, and a \texttt{pathlib}-based file iteration. This is Python that will read well in a pull request.
\end{winbox}

\begin{tipbox}[The Python to Unix bridge]
You already know \texttt{awk '\{print \$1\}'} extracts column 1. The Python equivalent is \texttt{line.split('\textbackslash t')[0]}. Your Unix skills transfer directly. When in doubt, prototype in awk, then scale up in Python once the logic has more steps.
\end{tipbox}


\section{Block 4: Pandas, Plotting, and Bio Libraries}

\sprint{5}{Pandas for Tabular Genomic Data}{25}
\begin{itemize}[nosep]
  \item Load GWAS summary statistics: \texttt{df = pd.read\_csv('sumstats.tsv', sep='\textbackslash t', dtype=\{'CHR': str\})}. Explicit \texttt{dtype} for chromosome prevents Pandas from silently parsing ``X'' as NaN.
  \item Filter: \texttt{sig = df[df['P'] < 5e-8]}, the genome-wide significant hits.
  \item Group by: \texttt{df.groupby('CHR')['P'].min()}, the best p-value per chromosome.
  \item Merge: \texttt{pd.merge(genotypes, phenotypes, on='SAMPLE\_ID')}, the most common GWAS join.
\end{itemize}
\begin{tipbox}[If you do not have a summary-stats file yet]
Download one from the GWAS Catalog (\url{https://www.ebi.ac.uk/gwas/}) or the PGS Catalog (\url{https://www.pgscatalog.org}). If you just want to practice the mechanics, fabricate a minimal file with \texttt{numpy.random} and a few thousand rows; the commands above work identically.
\end{tipbox}
\begin{winbox}[Checkpoint]
You loaded, filtered, and grouped a summary-statistics table. Compare \texttt{df.shape} before and after filtering to confirm the filter did something.
\end{winbox}

\sprint{6}{Visualization}{25}
\begin{itemize}[nosep]
  \item Manhattan plot: scatter of $-\log_{10}(P)$ vs.\ genomic position, colored by chromosome.
  \item QQ plot: expected vs.\ observed $-\log_{10}(P)$; the standard GWAS diagnostic for inflation.
  \item Use matplotlib for control and seaborn for defaults. Always label axes and include a title.
  \item Save with \texttt{plt.savefig('manhattan.png', dpi=150, bbox\_inches='tight')}.
\end{itemize}
\begin{winbox}[Checkpoint]
You produced a Manhattan plot and a QQ plot as PNGs. Commit both to your repo.
\end{winbox}

\sprint{7}{NumPy and SciPy for Fast Math}{25}
\begin{itemize}[nosep]
  \item NumPy arrays are typically two orders of magnitude faster than Python lists for numeric math. Use \texttt{np.mean(quality\_scores)} rather than \texttt{sum(xs)/len(xs)}.
  \item Vectorized operations avoid explicit loops: \texttt{genotypes * 2} multiplies every element at once.
  \item SciPy provides standard hypothesis tests: \texttt{scipy.stats.ttest\_ind(group1, group2)}.
  \item Multiple testing correction: \texttt{from statsmodels.stats.multitest import multipletests}.
\end{itemize}
\begin{winbox}[Checkpoint]
You ran a t-test in Python, then applied Benjamini--Hochberg correction to a vector of p-values and compared the number of significant hits before and after.
\end{winbox}

\sprint{8}{Bio Libraries: pysam, Biopython, scanpy}{25}
\begin{itemize}[nosep]
  \item pysam: read BAM/CRAM files directly. Count reads mapping to a gene interval in a few lines.
  \item Biopython: \texttt{SeqIO.parse()} handles FASTA, FASTQ, and GenBank; \texttt{Entrez.efetch()} queries NCBI.
  \item scanpy: the single-cell analysis framework used heavily in Blocks 9--10.
  \item scikit-allel: population genetics in Python (allele frequencies, LD, PCA from VCF data).
\end{itemize}
\begin{winbox}[Checkpoint]
You queried NCBI from Python (e.g., retrieved a GenBank record by accession) and counted reads in a BAM file for one interval.
\end{winbox}

\begin{restbox}[End of Blocks 3--4]
You just picked up a new language and applied it to biology. Take a day off. Let it consolidate.
\end{restbox}

\subsection{Checkpoint:}
\begin{itemize}[nosep]
  \item You can parse any common genomic file in Python (FASTA, BED, VCF, TSV)
  \item You can use Pandas to load, filter, group, and merge tabular data
  \item You can make a Manhattan plot and a QQ plot
  \item You understand why vectorized NumPy operations are faster than Python loops
\end{itemize}

\subsection{Resources}
\begin{itemize}[nosep]
  \item McKinney -- \textit{Python for Data Analysis} (the Pandas bible)
  \item VanderPlas -- \textit{Python Data Science Handbook} (free online)
  \item Real Python (realpython.com) --- searchable tutorials for anything
\end{itemize}
