\chapter{Blocks 5--6: R, Bioconductor \& Differential Expression}\label{chap:r}

\textbf{Primary:} Bioconductor, differential expression, ggplot2\quad|\quad\textbf{Interleave:} Statistics concepts, Unix file prep

\begin{brainbox}[Why R when you just learned Python?]
R has Bioconductor, a collection of thousands of biology-specific packages with no direct Python equivalent. DESeq2, edgeR, GenomicRanges, and Seurat all live there. For working genomicists, R and Python are complements rather than substitutes.
\end{brainbox}

\begin{tipbox}[Install R and RStudio first]
R is not installed by default on macOS or most Linux distributions. Download the pre-built binary for your OS/architecture from CRAN (\url{https://cran.r-project.org}), then install RStudio Desktop for a comfortable IDE (\url{https://posit.co/download/rstudio-desktop/}). See \hyperref[app:tools]{Appendix~A} for more.
\end{tipbox}

\begin{tipbox}[Data files for this chapter]
Sprints 2--5 read four small data files from the \texttt{data/} subdirectory of this curriculum's repo: \texttt{counts.tsv}, \texttt{sample\_info.tsv}, \texttt{de\_results.csv}, and \texttt{peaks.bed}. Either clone the repo and \texttt{setwd()} into it, or copy the files into the working directory of your R session. \texttt{file.exists("data/counts.tsv")} should return \texttt{TRUE} before you start Sprint 2.
\end{tipbox}


\section{Block 5: R Basics + Bioconductor}

\sprint{1}{R in 25 Minutes: Vectors, Lists, Data Frames}{25}
\begin{itemize}[nosep]
  \item \textbf{Vectors.} \texttt{x <- c(1, 2, 3); mean(x)}. Everything in R is a vector; even a single number is a length-1 vector. Vectors are homogeneous: one type throughout (numeric, character, or logical).
  \item \textbf{Lists.} Heterogeneous containers:
    \begin{lstlisting}
res <- list(pval = 0.003,
            estimate = 1.42,
            samples = c("A", "B", "C"))
res$pval          # 0.003
res[["samples"]]  # c("A", "B", "C")
    \end{lstlisting}
    Access elements by name with \texttt{\$} or by name/position with \texttt{[[}. Every complex R object (\texttt{lm()} output, a \texttt{DESeqDataSet}, a phylo tree) is a list underneath. Learn \texttt{str(obj)} now; you will use it constantly to inspect unfamiliar objects.
  \item \textbf{Data frames.} A list of equal-length columns with row-and-column indexing: \texttt{df <- read.csv('samples.tsv', sep='\textbackslash t')}. First three commands on any new data frame: \texttt{head(df)}, \texttt{names(df)}, \texttt{str(df)}.
  \item \textbf{Indexing gotchas.} R is 1-based: \texttt{df[1,]} is row 1, not row 2. Negative indices exclude: \texttt{df[-1,]} drops row 1. Logical vectors filter: \texttt{df[df\$pval < 0.05, ]}.
\end{itemize}
\begin{tipbox}[R quirks worth naming up front]
\textbf{Assignment operator: \texttt{<-} vs.\ \texttt{=}.} The R convention is \texttt{<-} for creating variables (\texttt{x <- 5}). The \texttt{=} sign is also accepted at the top level and does the same thing, but inside a function call it does something \textit{different}: it names an argument rather than creating a variable. Consider:
\begin{lstlisting}
mean(x = c(1, 2, NA), na.rm = TRUE)  # names the first and
                                     # second arguments
mean(x <- c(1, 2, NA), na.rm = TRUE) # creates a variable x
                                     # in the surrounding
                                     # scope AND passes it
                                     # positionally
\end{lstlisting}
The first form is what you want almost every time. The second form has a side effect (a new \texttt{x} visible outside the call) which is almost never what you want and is a classic source of confusing bugs. So: use \texttt{=} to match arguments inside \texttt{f(arg = value)} calls, and use \texttt{<-} to create variables at the top level. In practice, if you follow that rule both sides of your code read correctly and you never have to think about it again.

Does it really matter? Mostly for readability and for avoiding the side-effect bug above --- execution-wise, \texttt{x <- 5} and \texttt{x = 5} at the top level produce the same bindings. Most R style guides (tidyverse, Bioconductor, Google's) specify \texttt{<-} for assignment, so code you read in the wild will look that way. Editors like RStudio bind \texttt{Alt--} (Windows/Linux) or \texttt{Option--} (macOS) to type \texttt{<-} in one keystroke.

\textbf{\texttt{NA}} means ``missing'' and \textit{propagates} through arithmetic: \texttt{mean(c(1,2,NA))} returns \texttt{NA} unless you pass \texttt{na.rm = TRUE}. \textbf{Factors} look like strings but behave like integers; if a column's contents surprise you, run \texttt{class()} first.
\end{tipbox}
\begin{winbox}[Checkpoint]
You loaded a TSV, inspected its structure with \texttt{str()}, pulled one column and one row, and can describe the difference between a vector, a list, and a data frame. Apply functions and the tidyverse come next.
\end{winbox}

\sprint{2}{Lists, the Apply Family, and Functional R}{25}
R grew up as a functional language: instead of writing for-loops, you hand a function to another function that runs it over a collection. This pattern is everywhere in R code, so you need to read it fluently even if you prefer to write loops yourself.

\textbf{Load the count table for this sprint first.} \texttt{counts.tsv} has 50 genes and four count columns (\texttt{sample1}--\texttt{sample4}):
\begin{lstlisting}
counts_df <- read.table("data/counts.tsv",
                        header = TRUE, sep = "\t")
head(counts_df)
counts_mat <- as.matrix(
  counts_df[, c("sample1", "sample2",
                "sample3", "sample4")])
rownames(counts_mat) <- counts_df$gene_id
\end{lstlisting}
Now you have \texttt{counts\_mat}: 50 rows (genes) by 4 columns (samples), with gene names as row names.

\begin{itemize}[nosep]
  \item \textbf{\texttt{apply()} --- walk rows or columns of a matrix.} First arg = matrix; \texttt{1} = rows, \texttt{2} = columns; third arg = the function.
    \begin{lstlisting}
row_means <- apply(counts_mat, 1, mean)
col_sds   <- apply(counts_mat, 2, sd)
head(sort(row_means, decreasing = TRUE))
col_sds
    \end{lstlisting}
    \texttt{row\_means} is a per-gene mean count; \texttt{col\_sds} is a per-sample standard deviation.
  \item \textbf{\texttt{lapply()} --- run f over a list, return a list.} Treat the columns of \texttt{counts\_mat} as a list of vectors:
    \begin{lstlisting}
per_sample <- as.list(as.data.frame(counts_mat))
means <- lapply(per_sample, mean)
means$sample1
    \end{lstlisting}
  \item \textbf{\texttt{sapply()}} does the same and simplifies to a vector/matrix when it can: \texttt{sapply(per\_sample, mean)} returns a named numeric vector. Convenient interactively; prefer \texttt{vapply()} in scripts --- it type-checks the return shape and fails loudly when a single element deviates.
  \item \textbf{\texttt{tapply()} --- split by a grouping vector, then apply.} Use the \texttt{chromosome} column to compute mean expression per chromosome for sample1:
    \begin{lstlisting}
tapply(counts_df$sample1,
       counts_df$chromosome, mean)
    \end{lstlisting}
  \item \textbf{\texttt{mapply()} / \texttt{Map()} --- over multiple parallel inputs.} \texttt{mapply(t.test, group\_a\_list, group\_b\_list)} runs a t-test pairing element 1 of each, element 2 of each, etc.
  \item \textbf{Why this matters.} DESeq2 results come back as a list; \texttt{apply()} is how you compute per-gene statistics; \texttt{lapply()} appears in every production R pipeline. Loops are not wrong, but apply-family code is shorter and more idiomatic.
  \item \textbf{The modern equivalent.} The \texttt{purrr} package (part of the tidyverse) gives you \texttt{map()}, \texttt{map\_dbl()}, \texttt{map\_chr()}, \texttt{map2()}, \texttt{pmap()} --- same idea, stricter return types. We'll use \texttt{purrr::map\_dbl()} in the next sprint.
\end{itemize}
\begin{winbox}[Checkpoint]
You loaded \texttt{counts.tsv}, computed per-gene means and per-sample standard deviations, and got mean expression per chromosome with one \texttt{tapply} call --- without writing a single explicit loop. This is the R way.
\end{winbox}

\sprint{3}{Tidyverse: A Modern Layer on Top of R}{25}
Base R works, but routine data wrangling is verbose. The tidyverse is a family of packages (\texttt{dplyr}, \texttt{tidyr}, \texttt{readr}, \texttt{purrr}, \texttt{ggplot2}, \texttt{stringr}, \texttt{forcats}, and more) that share a consistent grammar: pipe a data frame through a sequence of \textit{verbs}, each returning a modified data frame. Once you internalize the verbs, most analyses become one-glance readable.

Install once with \texttt{install.packages("tidyverse")} in an R session --- that single call pulls in every package listed above. Docs: \url{https://www.tidyverse.org}.

\begin{itemize}[nosep]
  \item \textbf{The pipe.} \texttt{\%>\%} (from \texttt{magrittr}, re-exported by dplyr) and the newer base-R \texttt{|>} (R $\geq$ 4.1) both do the same thing: pass the left-hand side as the first argument of the right-hand side. They're Unix \texttt{|} for data frames.
  \item \textbf{Load the example DE table.} \texttt{de\_results.csv} has columns \texttt{gene}, \texttt{chromosome}, \texttt{sample}, \texttt{log2FC}, \texttt{p.value}, and \texttt{pAdj}. The dot in \texttt{p.value} is just part of the column name --- back-tick it when dplyr needs to quote it.
    \begin{lstlisting}
library(readr)
library(dplyr)
de_results <- read_csv("data/de_results.csv")
glimpse(de_results)
    \end{lstlisting}
  \item \textbf{The five verbs you'll use constantly.}
    \begin{lstlisting}
de_results |>
  filter(pAdj < 0.05) |>
  mutate(direction = if_else(log2FC > 0,
                             "up", "down")) |>
  group_by(chromosome, direction) |>
  summarise(n = n(), .groups = "drop") |>
  arrange(desc(n))
    \end{lstlisting}
    \texttt{filter} picks rows; \texttt{mutate} adds/changes columns; \texttt{group\_by} + \texttt{summarise} compute per-group statistics; \texttt{arrange} sorts; \texttt{select} picks columns. Five verbs cover 80\% of real analysis.
  \item \textbf{Tibbles.} \texttt{tibble()} creates a modern data frame that prints neatly (no auto-factor conversion, no surprising type coercion). \texttt{readr::read\_csv()} returns a tibble; it behaves like a \texttt{data.frame} everywhere else.
  \item \textbf{Reshaping.} \texttt{pivot\_longer()} and \texttt{pivot\_wider()} (\texttt{tidyr}) go between wide and long formats --- critical for plotting in ggplot2 and for joining disparate datasets.
  \item \textbf{Joins.} \texttt{inner\_join()}, \texttt{left\_join()}, \texttt{anti\_join()}. Merge GWAS summary stats with gene annotations by SNP ID in one line.
  \item \textbf{\texttt{across()} for column-wise operations.} \texttt{mutate(across(starts\_with("sample\_"), log1p))} log-transforms every sample column at once.
\end{itemize}
\begin{tipbox}[When to stay in base R]
Tidyverse is great for interactive analysis and scripts humans will read. Package authors often prefer base R internally --- it has zero dependencies and a more stable API. Both styles are valid; most real projects mix them.
\end{tipbox}
\begin{winbox}[Checkpoint]
You chained five dplyr verbs to go from raw DE results to a ranked table of up/down counts per chromosome. Compare the equivalent base-R code (\texttt{aggregate}, nested \texttt{[}) --- same output, very different ergonomics.
\end{winbox}

\sprint{4}{Bioconductor: GRanges for Genomic Intervals}{25}
\begin{itemize}[nosep]
  \item Install (one-time):
    \begin{lstlisting}
install.packages("BiocManager")
BiocManager::install(c("GenomicRanges",
                       "rtracklayer"))
    \end{lstlisting}
  \item \textbf{Create a GRanges by hand.}
    \begin{lstlisting}
library(GenomicRanges)
gr <- GRanges(seqnames = "chrI",
              ranges = IRanges(start = 100,
                               end   = 200))
gr
    \end{lstlisting}
  \item \textbf{Load real intervals from a BED file.} \texttt{peaks.bed} is the BED5 file (chrom/start/end/name/score) of 30 ChIP-style peaks on yeast chromosomes:
    \begin{lstlisting}
library(rtracklayer)
peaks <- import("data/peaks.bed")
peaks       # a GRanges with 30 ranges
seqnames(peaks)
width(peaks)
    \end{lstlisting}
    \texttt{rtracklayer::import()} reads BED, GFF, GTF, and BigWig with the same call.
  \item \textbf{Overlap two GRanges.} Make a few fake gene intervals and find which peaks fall in genes:
    \begin{lstlisting}
genes <- GRanges(
  seqnames = c("chrI", "chrII", "chrIV"),
  ranges   = IRanges(start = c(1, 100000, 800000),
                     end   = c(60000, 200000, 900000)))
hits <- findOverlaps(peaks, genes)
hits
    \end{lstlisting}
    \texttt{findOverlaps()} is the in-R counterpart of \texttt{bedtools intersect}.
\end{itemize}
\begin{winbox}[Checkpoint]
You imported a real BED file as a \texttt{GRanges} object and ran an overlap against a hand-built set of gene intervals. Same concept as bedtools, different syntax.
\end{winbox}

\sprint{5}{SummarizedExperiment: The Standard Bioconductor Container}{25}
\begin{itemize}[nosep]
  \item \textbf{Build one from the files you already have.} A \texttt{SummarizedExperiment} bundles three things into a single object: a counts matrix, gene-level metadata (\texttt{rowData}), and sample-level metadata (\texttt{colData}). Construct it from \texttt{counts.tsv} and \texttt{sample\_info.tsv}:
    \begin{lstlisting}
library(SummarizedExperiment)

counts_df <- read.table("data/counts.tsv",
                        header = TRUE, sep = "\t")
sample_df <- read.table("data/sample_info.tsv",
                        header = TRUE, sep = "\t")

count_mat <- as.matrix(
  counts_df[, c("sample1", "sample2",
                "sample3", "sample4")])
rownames(count_mat) <- counts_df$gene_id

se <- SummarizedExperiment(
  assays  = list(counts = count_mat),
  rowData = DataFrame(chromosome = counts_df$chromosome),
  colData = DataFrame(sample_df,
                      row.names = sample_df$sample))
se
    \end{lstlisting}
  \item \textbf{Explore the object you just built.}
    \begin{lstlisting}
dim(se)              # genes x samples
assay(se)[1:5, 1:4]  # top-left of count matrix
rowData(se)          # per-gene metadata
colData(se)          # per-sample metadata
    \end{lstlisting}
  \item \textbf{Why this matters.} Most Bioconductor analysis tools (DESeq2, edgeR, limma) accept a \texttt{SummarizedExperiment} directly. Bundling assay + rowData + colData in one object means you can subset by sample (\texttt{se[, se\$condition == "treated"]}) without ever desynchronizing the count matrix from its metadata --- a classic source of bugs when you keep them in parallel objects.
  \item \textbf{Optional: pull a public dataset.} For a larger real-world example, try \texttt{recount3} (pre-aligned RNA-seq from thousands of public studies) or \texttt{GEOquery} (any GEO series); both return \texttt{SummarizedExperiment} objects with the same accessor functions.
\end{itemize}
\begin{winbox}[Checkpoint]
You built a \texttt{SummarizedExperiment} from \texttt{counts.tsv} + \texttt{sample\_info.tsv}, and \texttt{dim(se)}, \texttt{colData(se)}, and \texttt{rowData(se)} all return something sensible. This is the object DESeq2 expects in the next sprint.
\end{winbox}


\section{Block 6: Differential Expression and ggplot2}

\sprint{6}{DESeq2: Standard RNA-seq Differential Expression}{50}
\begin{itemize}[nosep]
  \item Build a \texttt{DESeqDataSet} from your \texttt{SummarizedExperiment} plus a design formula (e.g., \texttt{\textasciitilde condition}).
  \item Run \texttt{dds <- DESeq(dds); res <- results(dds)}. Two function calls do the whole analysis.
  \item Understand what happened under the hood: normalization (size factors), dispersion estimation with empirical Bayes shrinkage, a negative-binomial GLM fit per gene, and a Wald test for each contrast.
  \item Extract results: \texttt{res[order(res\$padj), ]} gives genes sorted by adjusted p-value. Save the table to disk --- you'll plot it with ggplot2 in Sprint 7.
\end{itemize}
\begin{winbox}[Checkpoint]
You ran a differential expression analysis end-to-end. Sanity checks: a mostly-uniform p-value histogram away from zero, and a few genes with very low \texttt{padj}.
\end{winbox}

\sprint{7}{ggplot2: Grammar of Graphics}{25}
\begin{itemize}[nosep]
  \item The basic form: \texttt{ggplot(data, aes(x, y)) + geom\_point() + theme\_minimal()}.
  \item Layer more geoms: \texttt{+ geom\_violin() + geom\_boxplot(width = 0.1)}.
  \item Facets: \texttt{+ facet\_wrap(\textasciitilde chromosome)} splits into small multiples by a variable.
  \item Save with \texttt{ggsave('figure1.pdf', width = 8, height = 6)}.
  \item \textbf{Volcano plot.} Use the DE results from Sprint 6 (or \texttt{de\_results.csv} from Sprint 3) to make a volcano plot of $-\log_{10}(p)$ vs.\ $\log_2(\text{fold change})$:
    \begin{lstlisting}
library(ggplot2)
library(readr)
de <- read_csv("data/de_results.csv")
ggplot(de, aes(x = log2FC,
               y = -log10(p.value))) +
  geom_point(aes(color = pAdj < 0.05),
             alpha = 0.7) +
  scale_color_manual(values = c("grey60",
                                "firebrick")) +
  labs(x = "log2 fold change",
       y = "-log10(p-value)",
       color = "FDR < 0.05") +
  theme_minimal()
ggsave("volcano.pdf", width = 6, height = 5)
    \end{lstlisting}
    A well-behaved dataset shows a roughly symmetric volcano with the most extreme points highlighted.
\end{itemize}
\begin{winbox}[Checkpoint]
You produced a multi-panel figure with labeled axes, a title, and a saved PDF in your repo, plus a volcano plot from your DE results.
\end{winbox}

\sprint{8}{Compare DE Methods}{25}
\begin{itemize}[nosep]
  \item Run the same dataset through \texttt{limma-voom}: \texttt{voom(counts)} $\to$ \texttt{lmFit} $\to$ \texttt{eBayes} $\to$ \texttt{topTable}.
  \item Compare the result tables: how many genes are significant at FDR 0.05 in DESeq2 but not limma? Vice versa?
  \item Visualize the overlap with a Venn diagram or UpSet plot.
  \item Consider why they differ: DESeq2 fits a negative-binomial GLM; limma-voom transforms counts to approximately normal and fits a linear model with precision weights. Which assumption is more plausible for your data?
\end{itemize}
\begin{winbox}[Checkpoint]
You compared two methods on the same data, quantified the overlap, and articulated one reason the results might disagree.
\end{winbox}

\begin{restbox}[End of Blocks 5--6]
You now have working fluency in bash, Python, and R, and can run a differential expression analysis end-to-end. Take a break before moving on.
\end{restbox}

\subsection{Checkpoint:}
\begin{itemize}[nosep]
  \item You can manipulate data in R with tidyverse
  \item You understand GRanges and SummarizedExperiment
  \item You can run a DESeq2 analysis end-to-end
  \item You can make publication-quality ggplot2 figures
  \item You can explain what empirical Bayes shrinkage does and why DESeq2 uses it
\end{itemize}

\subsection{Resources}
\begin{itemize}[nosep]
  \item R for Data Science, 2nd Ed (r4ds.hadley.nz) --- free online, excellent
  \item DESeq2 vignette (follow the Beginner's Guide first, then the full vignette)
  \item StatQuest YouTube: the DESeq2 and edgeR episodes are outstanding
  \item Bioconductor workflows: bioconductor.org/packages/release/workflows/
\end{itemize}
