\chapter{Blocks 11--12: AI, Right Questions \& Capstone}\label{chap:capstone}

\textbf{Primary:} AI prompting, data expectations, reproducible pipelines, capstone project

\begin{brainbox}[The meta-skill]
Tools change. Methods change. The ability to ask the right question, recognize unexpected data, and evaluate AI output --- that lasts your whole career.
\end{brainbox}


\section{Block 11: Asking the Right Questions + AI Mastery}

\sprint{1}{The Question Framework}{25}
\begin{itemize}[nosep]
  \item Before every analysis, write down answers to: what is the biological question, what would a positive result look like, and what would a null result look like?
  \item Data inventory: how many rows? Columns? What does each represent? What are the expected ranges?
  \item Sanity checks: roughly 4--5 million variants per human genome, about 20{,}000 protein-coding genes, 500--5000 genes detected per cell in scRNA-seq.
  \item Red flags to watch for: a flat Manhattan plot (no signal at all), bimodal quality score distributions, PCs that correlate with batch instead of biology.
\end{itemize}
\begin{winbox}[Checkpoint]
You have a one-page pre-analysis checklist for the dataset you plan to work with. Commit it to your repo before you touch the data.
\end{winbox}

\subsection{The Question Map}

\begin{center}
\begin{tabularx}{\textwidth}{l X}
\toprule
\headertext{Stage} & \headertext{Ask This} \\
\midrule
Before you start & What is the question? What would a definitive answer look like? \\
Loading data & What do rows/columns represent? Expected dimensions? Missing values? \\
Choosing methods & What does this method assume? Does my data meet the assumptions? \\
After QC & Do distributions look right? Do controls work? Any batch effects? \\
Interpreting results & Is the effect size meaningful? Could confounders explain this? \\
Before publishing & Can someone else reproduce this? Are all parameters recorded? \\
\bottomrule
\end{tabularx}
\end{center}

\sprint{2}{Prompting AI Like a Bioinformatician}{25}
\begin{itemize}[nosep]
  \item Context first: ``I have a 10x scRNA-seq h5ad file, 50{,}000 cells, human PBMCs. I want to \ldots''
  \item Constraints: ``Use scanpy 1.9 or newer, avoid deprecated functions, include a comment at each QC threshold explaining the rationale.''
  \item Verification: ``Add assertions that check output dimensions and raise a clear error if something is unexpected.''
  \item Debugging: paste the full error message and traceback. Do not paraphrase; paste.
\end{itemize}
\begin{winbox}[Checkpoint]
You wrote a single prompt that produced working, well-commented code on the first run, and saved both the prompt and the output to your repo.
\end{winbox}

\sprint{3}{Spotting AI Hallucinations}{25}
\begin{itemize}[nosep]
  \item Common failure modes: inventing function names, mixing up package versions, using wrong parameter names, citing papers that do not exist.
  \item Always test on a small known-answer dataset before trusting the code on real data.
  \item Cross-check: ask two different AI tools the same question. If they disagree, dig in.
  \item Version trap: AI outputs frequently mix scanpy 1.8 vs.\ 1.10 syntax, or old Seurat v3 vs.\ v5 function names.
\end{itemize}
\begin{dobox}[Concrete exercise]
Ask an AI tool to ``write scanpy code that computes the mean expression of gene \textit{GAPDH} per Leiden cluster from an \texttt{AnnData} object.'' Run the code as-is. Then verify every function name and argument against the current scanpy documentation (\url{https://scanpy.readthedocs.io/}). If nothing is wrong, repeat the prompt with a deliberately obscure request (e.g., ``use \texttt{sc.tl.expression\_by\_cluster}'' --- this function does not exist). Note in your journal what the AI did when asked about an invented function.
\end{dobox}
\begin{winbox}[Checkpoint]
You have a journal entry documenting one verified, correct AI output and one case where the AI either invented a function or hallucinated confidently about an invented API you asked it to use.
\end{winbox}

\begin{warnbox}[AI ground rules for research]
\begin{itemize}[nosep]
  \item Always run AI-generated code before trusting it.
  \item Cite AI assistance in your methods section. Transparency matters.
  \item Never use AI for novel method development without deep verification.
  \item AI is a study partner, not a replacement for understanding.
\end{itemize}
\end{warnbox}


\section{Block 12: Reproducible Pipeline + Capstone}

\sprint{4}{Snakemake: Your Pipeline in Code}{50}
\begin{itemize}[nosep]
  \item Install: \texttt{mamba install -c bioconda -c conda-forge snakemake}. Docs: \url{https://snakemake.readthedocs.io/}.
  \item Snakemake defines a pipeline as a set of rules. Each rule has input files, output files, and a shell or Python command that turns one into the other.
  \item Start simple: a rule to download, a rule to run QC, a rule to align. Run with \texttt{snakemake --cores 4}.
  \item Snakemake figures out the execution order from the rules' inputs and outputs. If you change one input file, it re-runs only what depends on that file.
  \item Add a \texttt{conda:} directive to each rule to pin a per-step environment. At that point the pipeline is fully reproducible on any machine with Snakemake and conda.
  \item Alternative: Nextflow plus nf-core if you prefer. Install Nextflow with \texttt{curl -s https://get.nextflow.io | bash}. nf-co.re (\url{https://nf-co.re}) hosts community-curated pipelines for most common assays.
\end{itemize}
\begin{winbox}[Checkpoint]
You have a multi-step Snakemake pipeline in a Git repo that someone else could clone and run. Verify by changing one input file and confirming only the downstream rules re-run.
\end{winbox}

\subsection{Capstone Project}

This is your final project. It ties together everything from the past 11 blocks into a single, complete analysis.

\begin{dobox}[What to build]
\begin{itemize}[nosep]
  \item Choose a biological question and a public dataset (GEO, SRA, GWAS Catalog).
  \item Build a reproducible pipeline (Snakemake or Nextflow) in a Git repository.
  \item Analyze with Python and/or R, using appropriate statistical methods.
  \item Produce at least 3 publication-quality figures with captions.
  \item Write a 5--10 page report: question, methods, results, interpretation.
  \item Include an AI journal: how you used AI tools, what worked, what didn't.
\end{itemize}
\end{dobox}

\begin{tipbox}[Suggested capstone topics]
\begin{itemize}[nosep]
  \item Complete \hyperref[ps:4]{Problem Set 4 (QTL Mapping)} from the \hyperref[chap:running]{Running Project} --- this is the recommended capstone if you've been working through Problem Sets \hyperref[ps:1]{1}--\hyperref[ps:3]{3}
  \item Reproduce a published scRNA-seq analysis with updated methods (new normalization, better clustering)
  \item GWAS-to-function: fine-map a locus, colocalize with eQTLs, identify candidate genes
  \item Cross-ancestry PRS: build polygenic risk scores, test portability across populations
  \item CRISPR screen reanalysis: apply MAGeCK + BAGEL2 to a published dataset, compare hit lists
\end{itemize}
\end{tipbox}

\sprint{5}{The Reproducibility Test}{25}
\begin{itemize}[nosep]
  \item Clone your repo to a different computer, or ask a colleague or an AI agent to try.
  \item Can they install the environment and run the pipeline end-to-end from the commands in your README?
  \item If yes, you are done. If no, fix whatever broke and repeat.
\end{itemize}
\begin{winbox}[Checkpoint]
Someone else reproduced your analysis from your repo without asking you for help. That is the standard to aim for.
\end{winbox}

\begin{restbox}[End of the curriculum]
At this point you can process genomic data in Unix, analyze it in Python and R, apply rigorous statistics, and produce reproducible research. The foundation is in place; continued learning comes from applying these skills to real research questions.
\end{restbox}
